Neural Connectome of the Ctenophore Statocyst

### 2025-06-24 manuscript/Manuscript.qmd
MS and fig: change bristle to large granular cell

![**Figure 2. Organisation of the aboral synaptic nerve net**\
(A) Morphological rendering of the large aboral nerve-net neuron ANN_Q1-4 spanning all four quadrants, in aboral view (left), sagittal view (middle panel) and tentacular view (right panel). Spheres indicate the positions of nuclei.\
(B) Morphological rendering of the two smaller ANN neurons ANN_Q1Q2 (pink) and ANN_Q3Q4 (orange), each spanning two quadrants, in aboral view (left), sagittal view (middle panel) and tentacular view (right panel). Spheres indicate the positions of nuclei.\
(C) Number of mitochondria per cell for each cell type. The number of synapse-associated mitochondria is shown in green, the number of mitochondia outside synapses is shown in grey. N \> 2 cells for each cell type except lithocytes. Abbreviations: bal, balancer; brg, bridge; bsl, bristle; cg, ciliated groove; dv, dense vesicle cells; imc, intra-multiciliated cells; la, lamellate bodies; li, lithocytes; pl, plumose; ef, epithelial floor cells.\
(D) Position of mitochondria within ANN_Q1-4. Red indicates mitochondria associated with the presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.\
(E) Representative electron micrographs of presynaptic triad structures observed in the dataset and their schematic diagrams. The left diagram illustrates synaptic projections from the center ANN (ANN Q1-4) to the lateral ANN (ANN Q3Q4), while the right diagram shows synaptic projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN Q1-4). Abbreviations: mi, mitochondrion; er, endoplasmic reticulum; sv, synaptic vesicles. Scale bar: 500 nm.\
(F) Position of synapses. Synapses from the center ANN to the lateral ANN are shown in magenta, while synapses from the lateral ANN to the center ANN are shown in cyan. Blue dots indicate the locations of autapses within ANN_Q1-4.](figures/Figure2.png)

@@

![**Figure 3. Synaptic connectivity of the balancer and and bridge cells with the ANN.**\
(A) Number of synaptic inputs from the ANNs (ANN_Q1-4: blue, ANN_Q1Q2: orange, ANN_Q3Q4: pink) to each cell type. Abbreviations: bal, balancer; brg, bridge; bsl, bristle; cg, ciliated groove; dv, dense vesicle cell; imc, intra-multiciliated cell; la, lamellate body; li, lithocyte; pl, plumose cell; ef, epithelial floor cell.\
(B) Morphological rendering of balancer cells, with each cell color-coded by quadrant: Q1 (yellow), Q2 (blue), Q3 (orange), and Q4 (magenta). Each sphere represents the position of an individual nucleus. The fine projections extending from the cells represent traced trajectories that follow the cell body and continue into a single cilium from the point where it emerges. Views are shown from three orthogonal perspectives: aboral view (left), sagittal plane (middle), and tentacular plane (right). The spatial arrangement highlights the quadrant-specific organization of balancer clusters within the aboral organ. Scale bar: 25 µm.\
(C) Morphological rendering of bridge cells spanning the Q1Q2 and Q3Q4 quadrants. The Q1Q2-side bridge cells (8 cells) are shown in blue, while the Q3Q4-side bridge cells (6 cells) are shown in xxx color. The morphology of individual bridge cells extending across opposite quadrant regions is depicted. The spheres represent the positions of individual nuclei. The left panel shows an aboral view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(D) Mitochondrial localization within bridge cells and associated presynaptic triad structures. Red indicates mitochondria associated with presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified. The left panel shows a dorsal view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(E) Synaptic connections from ANN neurons to balancer ciliated cells. The positions of synapses from ANNs to balancers (magenta) are indicated. The left panel presents an aboral view, while the right panel shows a lateral view and a tentacular plane view. Balancer ciliated cells are depicted in light gray.\
(F) Synaptic connections between ANNs and bridge cells. The positions of synapses from ANNs to bridge cells (magenta) and from bridge cells to ANNs (light blue) are indicated. The left panel shows an aboral view, while the right panel presents a tentacular plane lateral view. Bridge cells are shown in light gray.\
(G) Connectivity matrix of the gravity-sensing neural circuit. Columns represent presynaptic cell groups, while rows represent postsynaptic cell groups. The numbers and varying shades of blue correspond to the number of synapses.\
(H) Complete connectivity map of the gravity-sensing neural circuit. Cells belonging to the same group are enclosed in hexagons, and the number of cells is added to their labels. The thickness of the arrows and the numerical values indicate the number of synapses. ANN Q1-4 are shown in blue, ANN Q1Q2 in orange, ANNs in Q3Q4 in pink, balancer cell groups in gray, and bridge cell groups in yellow.](figures/Figure3.png)

@@

We assigned each cell a specific cell-type name based on its category (balancer, bridge, bristle, ciliated grooves, dense vesicle cell, dome, epithelial floor (ef), intracellular multiciliated cells (imc), lamellate bodies (lb), lithocyte, plumose cell, ANN) resulting in a total of 12 cell types. Some cells only had more general features and were classified based on the presence/absence and number of cilia into four ciliated cell types:(biciliated (biC), monociliated (monoC), multiciliated (multiC), non-ciliated (non-C)). Additionally, we appended the quadrant number to the cell type name after an underscore ("\_") to indicate the cell-body’s location. For example, cells located in the first quadrant are named with "\_Q1." If a cell is situated between the first and second quadrants, we appended "Q1Q2" to the cell-type name.

@@

![**Figure 1—figure supplement 1. Cell-type composition of the ctenophore aboral organ**\
Morphological rendering of skeletonised cell types, (A) balancer cells (B) bridge cells, (C) bristle cells, (D) ciliated groove cells, (E) dense vesicle cells, (F) dome cells, (G) intra-multiciliated cells, (H) lamellate bodies, (I) lithocytes, (J) plumose cells, (K) aboral nerve-net neurons (L) epithelial floor cells.](figures/Figure1_Supplement1.png)

>>>>

![**Figure 2. Organisation of the aboral synaptic nerve net**\
(A) Morphological rendering of the large aboral nerve-net neuron ANN_Q1-4 spanning all four quadrants, in aboral view (left), sagittal view (middle panel) and tentacular view (right panel). Spheres indicate the positions of nuclei.\
(B) Morphological rendering of the two smaller ANN neurons ANN_Q1Q2 (pink) and ANN_Q3Q4 (orange), each spanning two quadrants, in aboral view (left), sagittal view (middle panel) and tentacular view (right panel). Spheres indicate the positions of nuclei.\
(C) Number of mitochondria per cell for each cell type. The number of synapse-associated mitochondria is shown in green, the number of mitochondia outside synapses is shown in grey. N \> 2 cells for each cell type except lithocytes. Abbreviations: bal, balancer; brg, bridge; lgc, large granular cell; cg, ciliated groove; dv, dense vesicle cells; imc, intra-multiciliated cells; la, lamellate bodies; li, lithocytes; pl, plumose; ef, epithelial floor cells.\
(D) Position of mitochondria within ANN_Q1-4. Red indicates mitochondria associated with the presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.\
(E) Representative electron micrographs of presynaptic triad structures observed in the dataset and their schematic diagrams. The left diagram illustrates synaptic projections from the center ANN (ANN Q1-4) to the lateral ANN (ANN Q3Q4), while the right diagram shows synaptic projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN Q1-4). Abbreviations: mi, mitochondrion; er, endoplasmic reticulum; sv, synaptic vesicles. Scale bar: 500 nm.\
(F) Position of synapses. Synapses from the center ANN to the lateral ANN are shown in magenta, while synapses from the lateral ANN to the center ANN are shown in cyan. Blue dots indicate the locations of autapses within ANN_Q1-4.](figures/Figure2.png)

@@

![**Figure 3. Synaptic connectivity of the balancer and and bridge cells with the ANN.**\
(A) Number of synaptic inputs from the ANNs (ANN_Q1-4: blue, ANN_Q1Q2: orange, ANN_Q3Q4: pink) to each cell type. Abbreviations: bal, balancer; brg, bridge; lgc, large granular cell; cg, ciliated groove; dv, dense vesicle cell; imc, intra-multiciliated cell; la, lamellate body; li, lithocyte; pl, plumose cell; ef, epithelial floor cell.\
(B) Morphological rendering of balancer cells, with each cell color-coded by quadrant: Q1 (yellow), Q2 (blue), Q3 (orange), and Q4 (magenta). Each sphere represents the position of an individual nucleus. The fine projections extending from the cells represent traced trajectories that follow the cell body and continue into a single cilium from the point where it emerges. Views are shown from three orthogonal perspectives: aboral view (left), sagittal plane (middle), and tentacular plane (right). The spatial arrangement highlights the quadrant-specific organization of balancer clusters within the aboral organ. Scale bar: 25 µm.\
(C) Morphological rendering of bridge cells spanning the Q1Q2 and Q3Q4 quadrants. The Q1Q2-side bridge cells (8 cells) are shown in blue, while the Q3Q4-side bridge cells (6 cells) are shown in xxx color. The morphology of individual bridge cells extending across opposite quadrant regions is depicted. The spheres represent the positions of individual nuclei. The left panel shows an aboral view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(D) Mitochondrial localization within bridge cells and associated presynaptic triad structures. Red indicates mitochondria associated with presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified. The left panel shows a dorsal view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(E) Synaptic connections from ANN neurons to balancer ciliated cells. The positions of synapses from ANNs to balancers (magenta) are indicated. The left panel presents an aboral view, while the right panel shows a lateral view and a tentacular plane view. Balancer ciliated cells are depicted in light gray.\
(F) Synaptic connections between ANNs and bridge cells. The positions of synapses from ANNs to bridge cells (magenta) and from bridge cells to ANNs (light blue) are indicated. The left panel shows an aboral view, while the right panel presents a tentacular plane lateral view. Bridge cells are shown in light gray.\
(G) Connectivity matrix of the gravity-sensing neural circuit. Columns represent presynaptic cell groups, while rows represent postsynaptic cell groups. The numbers and varying shades of blue correspond to the number of synapses.\
(H) Complete connectivity map of the gravity-sensing neural circuit. Cells belonging to the same group are enclosed in hexagons, and the number of cells is added to their labels. The thickness of the arrows and the numerical values indicate the number of synapses. ANN Q1-4 are shown in blue, ANN Q1Q2 in orange, ANNs in Q3Q4 in pink, balancer cell groups in gray, and bridge cell groups in yellow.](figures/Figure3.png)

@@

We assigned each cell a specific cell-type name based on its category (balancer, bridge, large granular, ciliated grooves, dense vesicle cell, dome, epithelial floor (ef), intracellular multiciliated cells (imc), lamellate bodies (lb), lithocyte, plumose cell, ANN) resulting in a total of 12 cell types. Some cells only had more general features and were classified based on the presence/absence and number of cilia into four ciliated cell types:(biciliated (biC), monociliated (monoC), multiciliated (multiC), non-ciliated (non-C)). Additionally, we appended the quadrant number to the cell type name after an underscore ("\_") to indicate the cell-body’s location. For example, cells located in the first quadrant are named with "\_Q1." If a cell is situated between the first and second quadrants, we appended "Q1Q2" to the cell-type name.

@@

![**Figure 1—figure supplement 1. Cell-type composition of the ctenophore aboral organ**\
Morphological rendering of skeletonised cell types, (A) balancer cells (B) bridge cells, (C) large granular cells, (D) ciliated groove cells, (E) dense vesicle cells, (F) dome cells, (G) intra-multiciliated cells, (H) lamellate bodies, (I) lithocytes, (J) plumose cells, (K) aboral nerve-net neurons (L) epithelial floor cells.](figures/Figure1_Supplement1.png)
